A primer to numerical simulations: The perihelion motion of Mercury

### 2017-12-01 tex/MercuryNumerical.tex
upadeted orbit figures

@@

\begin{figure}[htb]
	\centering
	\includegraphics[width=.25\textwidth]{figs/MercuryOrbit.png} \quad
	\includegraphics[width=.273\textwidth]{figs/PerihelionMotionLargeV.png} \quad
	\includegraphics[width=.26\textwidth]{figs/PerihelionMotion.png}
	\todo[inline]{@CK: reproduce figure}
	\caption{\label{fig:MercuryOrbit} Output for $\alpha=0$ (left), $\alpha=10^5$ and bigger starting distance $r_{MS}(0)=6R_0$ (middle) and $\alpha=10^5$ with original starting values.}
\end{figure}

@@

To point the students to this problem, they can be asked to increase their time steps significantly and observe the effect on the trajectories.
For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
In figure \ref{fig: LargeDT} the mercury trajectory for $dt=2 * v_M(0) / c_a \cdot 5$ is shown.
One can clearly see the failure of the simulation to reproduce the physical trajectory.
From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
One could quantify the deviation of $\delta\Theta$ from zero and use this to estimate the numerical systematic uncertainty for a given time step $dt$.
Only in the limit $dt \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.
\begin{figure}[htb]
	\centering
	\includegraphics[width=.3\textwidth]{figs/TrajectoryLargeDT.png}
	\caption{\label{fig: LargeDT} Deviation of the trajectory from the analytic result (indicated in blue) for large time steps (here $dt=2 * v_M(0) / a_M \cdot 5$)}
\end{figure}

>>>>

\usepackage{subcaption} % subfigures

@@

\begin{figure}[htb]
	\centering
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\caption{\label{fig:MercuryOrbit}%
		Different mercury orbits for $\alpha=0$ (a), $\alpha = 10^{6}$ (b) and $\alpha = 10^{6}$ but $r_{MS}(0) = 6 R_0$ (c).
		All computations where executed with $dt = 2 * v_{M}(0) / c_a / 10$.
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "mercury year" changes.
	}
\end{figure}

@@

To point the students to this problem, they can be asked to increase their time steps significantly and observe the effect on the trajectories.
For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
In figure \ref{fig: LargeDT} the mercury trajectory for $dt=2 * v_M(0) / c_a \cdot 5$ is shown.
One can clearly see the failure of the simulation to reproduce the physical trajectory.
From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
One could quantify the deviation of $\delta\Theta$ from zero and use this to estimate the numerical systematic uncertainty for a given time step $dt$.
Only in the limit $dt \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.
\begin{figure}[htb]
	\centering
	\includegraphics[width=.5\textwidth]{figs/num-err.png}
	\caption{\label{fig: LargeDT}%
		Deviation of the trajectory from precise numeric result (black line; $dt=2 v_M(0) / a_M / 20$) for large time steps (red line; $dt=2 v_M(0) / a_M \times 2$).
		Both orbits are plotted for $T = 10 T_M$.
	}
\end{figure}
